Does differentiation beget differentiation?

### 2022-11-02 Manuscript_3_draft1.Rmd
Additions made to the appendix as well as well as rewrites of the main text.

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized on `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research. However, we still know do not know whether the increased familiarity with differentiated integration that exposure to it brings leads to greater support a differentiated EU, and whether the mode of implementation and attitudes towards the EU moderates this effect. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that familiarity with differentiation seems to breed greater support for it. This particularly applies to Eurosceptics. It also suggests that support for differentiated integration may be higher where governments frame it as a source of greater national autonomy. My results suggest that exposure to the differentiated integration used to integrate highly salient policy areas can solidify popular support for a less integrated EU. This can complicate the EU’s move towards ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: TRUE
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

@@

Despite the EU being one of the world's most advanced political cooperations, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty. The differentiation of EU integration has primarily been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called "constitutional differentiation". Second, "instrumental differentiation" has happened where countries with a diminished capacity for integration have been allowed to integrate at different schedules. Such differentiation can either happen voluntarily or because the EU has imposed differentiation as a precondition for membership [@schimmelfennig_instrumental_2014]. Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's political development than the EU's own goal of "ever closer Union".

While increasing attention is being paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023], we know little about how actual exposure to it has shaped support for the goal of a uniformly integrated EU. While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the perceived democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain. This article tests three theoretically informed hypotheses: First, it tests whether support for differentiation leads to greater support for it on average. Second, it tests whether those more critical of EU membership become more positively inclined to differentiated integration if exposed to more of it. Lastly, it tests whether those citizens exposed to differentiated integration that was voluntarily chosen by their governments are more favourably inclined towards future differentiation.

By combining data from surveys fielded in 2020-2021 that measure citizens' attitudes towards constitutional differentiated integration in 13 EU member states with data detailing every country's exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism. I also find that support for it seems to be greater where differentiated integration has been framed as a tool for strengthening national sovereignty. This shows that familiarity with differentiated integration can lead to greater support for it in the future. However, more research is needed to understand what causes regional variations in support for differentiated integration.

In addition to expanding our knowledge of public support for differentiated EU integration, this article contributes to a broader literature on public support for international organizations [@dellmuth_legitimacy_2020; @dellmuth_elite_2021; @edwards_public_2009]. While the EU's legal and functional reach goes beyond that of most other international organization, many international organizations have seen their legal and political powers expand in recent years [@zurn_theory_2018]. Even if more research is needed to understand whether my results travel to contexts beyond the EU, they help illuminate how allowing countries to opt out of international rule-making shapes attitudes towards the future organization of the international legal order.

My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: Because EU integration has expanded into increasingly salient areas, differentiation has often been used to reduce public contestation of integration [@schimmelfennig_ever_2020]. By providing greater numbers of citizens with exposure to differentiation, the EU's institutions may solidify opposition to further integration among greater numbers of citizens, potentially stalling future attempts at cooperation [@hooghe_postfunctionalist_2009]. In this sense, the EU's differentiated integration may be both a help and hindrance for future integration: On the one hand, it may help facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013]. On the other, greater popular support for differentiated modes of integration can make it more difficult for the EU's institutions to get the public's buy-in to functionally necessary integration. The inevitable trade-off that this entails must be reckoned with by EU policy-makers deciding whether to allow for opt-outs from policy integration or to insist upon legal uniformity.

@@

An under-investigated question is why some people support a more differentiated EU, which renders the EU's integration either permanently or temporarily non-uniform. Differentiated integration is already investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration have focused on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022]. Thus, with one notable exception, the effect of national contexts are under-studied. Such national contexts are crucial for explaining attitudes towards both the EU and other international organizations [@dellmuth_elite_2021; @edwards_public_2009; @harteveld_europe_2013]. Here citizens' exposure to past differentiated integration plays a crucial role: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would preferences. The same citizens may thus respond by changing their preferences [@busemeyer_beyond_2021; @wlezien_public_1995]. While there have been many studies of the feedback loop between public opinion and the EU's ordinary policy-making [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are in essence alterations to the EU's institutional structure.

It is reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration [@marquart_knowing_2019; @verhaegen_does_2015]. Similarly, familiarity with more differentiated ways of structuring EU memberships may lead to a stronger perception among citizens that EU memberships that allow countries to opt of policy integration may be less problematic.

This positive relationship is likely to be more strongly felt for those that oppose their country's EU membership. Since constitutional differentiated integration allows for permanent opt-outs from policy integration, Eurosceptics who favour leaving the EU may see differentiated integration as a potentially more realistic path towards the less integration Europe that they prefer. This yields two hypotheses:  
*H1a: Exposure to differentiated integration increases support for future differentiated integration.*  
*H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*

Responses to differentiated integration may also vary depending on the situation in which differentiated integration was imposed. It is for instance likely that citizens may become more favourably inclined towards differentiated integration if the form of differentiated integration they have been exposed to was a result of a conscious decision to opt out on the part of their own governments. The reason is that these decisions may be more likely to be seen by citizens as an exercise of their own national sovereignty within the remits of an EU membership.

Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015; @spendzharova_banking_2016], and has typically been decided as a result of a referendum. Previous literature has found that the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020]. This shows that Scandinavian citizens may theoretically respond to differentiated integration, because of the particularly voluntary nature of how differentiated integration has been implemented in these regions, by wanting more differentiated integration in the future.

Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour. This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty. I thus hypothesize:

@@

To analyse the regionally specific effects of differentiation, I create dummies for Scandinavian and Central and Eastern European citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy. These are used to investigate the interaction effects assumed by *H2*. I lastly create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0. While this only measures what @de_vries_euroscepticism_2018 terms "exit scepticism", an alternative operationalization that uses a measure of satisfaction with EU democracy yields substantively the same results (see Robustness checks).

@@

```{r}
library(modelsummary)
table_format <- ifelse(is_docx, "huxtable", 'default')
modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                 title = "Descriptive statistics", output = "latex")
```

@@

A further limitation, general to all studies using observational survey data, is the potential for omitted variable bias. Here omitted variable bias is particularly likely to stem from the absence of data on the respondents' knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019]. I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias. I show the results of these analyses in the section Robustness checks. Lastly, the observational data found in this survey cannot be used to make counter-factual claims in the vein of @pearl_causal_2010.

@@

```{r }
main_mod = list()
main_mod$"Euroscepticism" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                         eurosceptic + gender + income_scale + age + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, 
                                                         data = EUI_final_DI_disc)
coef_map_eurosceptic <- c('(Intercept)' = 'Constant',
                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
                          "article_disc_log:nordic_dummy1" = "Article-years (logged) X Scandinavia", 
                          "article_disc_log:CEE_dummy1" = "Article-years (logged) X CEE",
                          "nordic_dummy1" = "Scandinavia",
                          "CEE_dummy1" = "Central and Eastern Europe",
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'age' = 'Age',
                          'income_scale' = 'Income',
                          'excl_identity1' = 'Exclusively national identity')
modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
             stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI. Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

Both table 2 and figure 1 show that exposure to differentiated integration is associated with greater support for differentiated integration as a future model of integration, and that this effect is particularly prominent among Eurosceptics.

@@

A reasonable counter-argument is that governments are more likely to opt for differentiated integration where underlying popular scepticism of integration may make citizens more likely to accept the idea of differentiated integration. Figure A6 offers some support for this assumption, as there are particularly high levels of differentiation in for instance Sweden and Denmark. These are countries where opt-outs from integration have been deployed as a response to broad Euroscepticism [@leruth_operationalizing_2015; @spendzharova_banking_2016]. However, citizens of countries as diverse as Lithuania, Hungary, Italy and the Netherlands have also been exposed to differentiated integration. It is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identify as a Eurosceptic periphery. This suggests that the correlations do not solely reflect an endogenous relationship between Euroscepticism and differentiation.

@@

modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
             stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
             vcov = ~country, gof_map = c("nobs", "adj.r.squared"))

```

@@

```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI (A = Scandinavia and B = CEE). 95% country-clustered CIs. Fixed effects omitted due to multicollinearity.", fig.height=8.5}
library(patchwork)

@@

@@

I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several models to omitted variable bias (see figures A1-4). The test quantifies the variance that a confounding variable would need to account for to shrink the effect of exposure to differentiated integration to zero. A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify the effect of the variable. One important omitted variable is knowledge of the EU [@leuffen_public_2022]. While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, its strong connection to support for differentiation makes its omission particularly problematic. This shows that omitted variable bias is a real concern for this study. Future studies should mitigate this by explicitly measuring knowledge of EU politics, to allow researchers to include a control similar to that found in other EU public opinion research.

I also use multiple imputation [@buuren_mice_2011] to test how sensitive the results are to bias caused by potentially systematic data missingness. By pooling the estimates and standard errors from 25 data imputations according to "Rubin's rules" [@rubin_2004_multiple] I better account for the uncertainty associated with imputation of missing values. This is particularly beneficial when conducting significance test. The effects of the multiple imputation suggests that there may be reason to think that systematic missingness is at play: Table A2 suggests that the effect of the interaction between exposure to differentiation and Euroscepticism loses its significance when the data are imputed. Similarly, we find that there is no longer a significant interaction effect between exposure and Euroscepticism in Scandinavia (table A4). However, the interaction effect of exposure and Scandinavian citizenship appears strongly robust to imputation.

I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the very low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`, of the measure of exposure to differentiated integration found in the Scandinavian data. I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data. Appendix table A6 shows that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4. The results thus suggest that the estimates are not unduly biased by the small variability found in the Scandinavian data.

@@

This paper establishes, by combining data surveying citizen preferences for differentiated integration with information about their countries' actual exposure to differentiation, that there may be a strong association between support for differentiated integration and past exposure to it. This may be because citizens more accustomed to differentiated integration have become more favourably inclined towards it. Importantly, the effect appears moderated by whether one opposes or supports EU membership. It also seems to be the case that exposure to frames depicting differentiated integration as a source of greater autonomy could be an important driver of support for differentiated integration

The fact that even Eurosceptics become more favourably inclined to differentiated integration the more they are exposed to it, despite a continued membership of the EU being out of line with what they actually want, suggests that differentiated integration may be seen as an acceptable "second-best" alternative to an EU exit even among those considering themselves opponents of membership. This furthermore strengthens the assumption that differentiated integration, as has been pointed to in previous literature [@bellamy_republican_2019; @schimmelfennig_ever_2020], can serve to mitigate the contestation of European integration brought by the EU's expansion into more important areas.

>>>>

---
title: "Does differentiation beget differentiation?"
subtitle: Examining the link between implementation of and support for differentiation
thanks: "This is a very early draft. All comments are welcome at martin.moland@arena.uio.no. This version was finalized on `r format(Sys.time(), '%B %d, %Y')`."
author:
- name: Martin Moland
  affiliation: ARENA, Centre for European Studies
# bibliography: "`r paste0(Sys.getenv('HOME'),'/Dropbox/master.bib')`" # This would be more general
bibliography: inst/bib_paper_3.bib
csl: inst/european-journal-of-political-research.csl
abstract: "Questions about the public legitimacy of the EU’s flexible integration is emerging as an important question for European public opinion research. However, we still know do not know whether the increased familiarity with differentiated integration that exposure to it brings leads to greater support a differentiated EU, and whether the mode of implementation and attitudes towards the EU moderates this effect. Combining recent survey data with data showing country-level exposure to differentiated integration, I show that familiarity with differentiation seems to breed greater support for it. This especially applies to Eurosceptics. It also suggests that support for differentiated integration may be higher where governments frame it as a source of greater national autonomy. My results suggest that exposure to the differentiated integration used to integrate highly salient policy areas can solidify popular support for a less integrated EU. This can complicate the EU’s move towards ‘ever closer Union’."
keywords: "differentiated integration, public opinion, Euroscepticism, European integration, contestation of the EU"
geometry: margin=1in
mainfont: cochineal
#sansfont: Linux Biolinum O
fontsize: 12pt
endnote: no
# pandocparas: TRUE
sansitup: FALSE
params:
  anonymous: ""
  doublespacing: TRUE
  removetitleabstract: ""
anonymous: "`r params$anonymous`"
doublespacing: "`r params$doublespacing`"
removetitleabstract: "`r params$removetitleabstract`"
header-includes:
  - \usepackage{longtable}
  - \LTcapwidth=.95\textwidth
  - \linespread{1.05}
  - \usepackage{hyperref}
  - \usepackage{setspace}
  - \usepackage{setspace}\doublespacing
---

@@

Despite the EU being one of the world's most advanced political cooperations, its integration has become increasingly fragmented after the ratification of the 1992 Maastricht Treaty. This fragmentation has been driven by a desire to deepen political cooperation even where it may be contested by some of the EU's members [@leuffen_differentiated_2013; @schimmelfennig_ever_2020], and has typically taken two forms: First, some member states have opted out of cooperation to protect their own sovereignty. This is called "constitutional differentiation". Second, "instrumental differentiation" happens where countries with a diminished capacity for integration have been allowed to integrate at different schedules. Such differentiation can either happen voluntarily or because the EU has imposed differentiation as a precondition for membership [@schimmelfennig_instrumental_2014]. Thus, 'variable geography' [@stubb_categorization_1996] better describes the EU's current development than its own normative goal of "ever closer Union".

While increasing attention is paid to what shapes popular preferences for such differentiation [@leuffen_public_2022; @telle_elite-mass_2022; @schuessler_mapping_2023; @de_blok_blessing_2023], we know little about how actual exposure to it has contributed to changing attitudes towards differentiation. While one of the few existing contributions examining the effect of differentiation on public opinion shows that allowing for differentiation may strengthen the democratic legitimacy of the EU [@schraff_does_2020], important knowledge gaps remain. This article tests two theoretically informed hypotheses: First, it tests whether exposure to differentiation leads to greater support for it on average, especially where the non-uniformity is voluntarily chosen. Second, it tests whether this effect differs by already existing attitudes towards EU membership.

By combining data from surveys fielded in 13 EU member states in 2020-2021 measuring citizens' attitudes towards constitutional differentiated integration with data detailing their countries' exposure to differentiated integration, I find that support for constitutional differentiation is associated with greater exposure to differentiations from EU law, but that this effect is strongly contingent on individual-level Euroscepticism. I also find that support for it seems to be greater where differentiated integration has been framed as a tool for strengthening national sovereignty. This suggests both that familiarity with differentiated integration can lead to greater support for it in the future, and that elite discourses matter for forming attitudes towards it.

My findings are also important to the discussion about the EU's future spawned by Brexit [@european_commission_white_2017]: The results suggest that the EU's differentiated integration may facilitate future integration of political powers traditionally considered core to statehood [@genschel_differentiated_2013]. This is made particularly clear by how even those who oppose EU membership appear particularly supportive of differentiated integration. That even those critical of EU membership may be favourably inclined towards the general principle of European integration as long as it allows countries to exercise autonomy when faced with the prospect of particularly controversial integration is an important finding for policy-makers seeking a way forward for the post-Brexit EU.

However, greater exposure to differentiated modes of integration can also complicate the EU's future integration if it leads to greater opposition to today's normative goal of 'ever closer Union'. The inevitable trade-off that this entails must be reckoned with by EU policy-makers deciding whether to allow for opt-outs from policy integration or to insist upon legal uniformity.

@@

An under-investigated question is why some people support differentiated integration, that renders the EU's integration permanently or temporarily non-uniform. Differentiated integration is already investigated conceptually [@batora_towards_2020; @leuffen_differentiated_2013; @schimmelfennig_instrumental_2014], empirically [@ganzle_differentiated_2019; @malang_political_2020; @ott_between_2017; @schimmelfennig_eastern_2017; @winzen_government_2020] and normatively [@bellamy_republican_2019; @eriksen_contesting_2019; @lord_autonomy_2021; @nicolaidis_new_2004]. The few contributions investigating individual support for differentiated integration focus on liberal economic values [@leuffen_public_2022], Euroscepticism [@de_blok_blessing_2023; @schuessler_mapping_2023], territorially exclusive self-identification [@moland_opting_2022] and party cues [@telle_elite-mass_2022]. However, we still do not know what impact exposure to differentiated integration in the past has on support for it as a future model of EU integration. Knowing whether exposure to differentiated integration has shaped attitudes towards it is important for two reasons: First, it can tell us whether the EU's current strategy of using differentiated integration to facilitate deeper EU integration [@schimmelfennig_ever_2020] has the counter-intuitive effect of solidifying support for a less integrated EU. Secondly, understanding whether exposure to differentiated integration leads to varying degrees of acceptance of it among opponents or supporters of EU membership is important because it can show whether differentiated integration can be used as a tool for reducing the contestation of European integration among citizens who are particularly critical of the EU.

It is theoretically plausible that exposure to differentiated integration will shape public support for it: Differentiations from EU law may create a disequilibrium between a country's actual level of integration and what its citizens would prefer. The same citizens may thus respond by changing their preferences [@busemeyer_beyond_2021; @wlezien_public_1995]. While the feedback loop between public opinion and the EU's ordinary policy-making has been extensively studied [@bolstad_dynamics_2015; @de_bruycker_democratically_2020], we do not know how this feedback loop plays out when the policies being made are alterations to the EU's institutional framework.

It is also reasonable to assume that the greater familiarity with differentiated integration that exposure to it will bring will lead to greater support for it in the future: Knowledge of EU policies is often, though not always, associated with greater support for EU integration, even if higher levels of general knowledge of EU politics do not always translate to either higher levels of knowledge of specific policies [@marquart_knowing_2019; @karp_know_2003; @verhaegen_does_2015]. As @karp_know_2003 point out, the theoretical mechanism that we would expect to be at play in this case is that greater familiarity with the EU and its policies would lead citizens to perceive EU integration as less of a threat than if they had no knowledge of the policy in question. Thus, someone who supports EU membership without having been exposed to differentiated integration might find what can only be perceived as a deviation from the uniformity that is frequently touted as a normative goal of the EU's integration problematic, and wonder how it would impact their membership of the EU. However, someone familiar with differentiated integration might be less worried about how such differentiation could impact their countries, as they have first-hand experience with how differentiated integration can combine both greater autonomy with a similar level of influence within the EU as what they had before the implementation of differentiated integration.

This positive relationship is likely to be more strongly felt for those that oppose their country's EU membership. For them, the perception that differentiated integration can better combine EU membership with national autonomy may be particularly salient, because they would be particularly likely to favour modes of integration that preserve a sovereignty they are particularly likely to believe to be important. This yields two hypotheses:  
*H1a: Exposure to differentiated integration increases support for future differentiated integration.*  
*H1b: Exposure will lead to greater support for differentiation among Eurosceptics than supporters of EU membership.*

It is also theoretically reasonable to assume that citizens may become more favourably inclined towards differentiated integration if the differentiation was brought about as a result of a conscious decision to opt out of policy integration. This would especially be the case if the opt-out was brought about through political debate, as this would both make the issue of opt-outs salient for individual citizens and make it more likely that citizens would be exposed to discourses that frame differentiated integration as a tool for exercising greater autonomy.

Scandinavia offers an interesting case study of this assumption: Here differentiation has been chosen by national governments and framed as a tool for strengthening national sovereignty [@adler-nissen_opting_2014; @leruth_operationalizing_2015], and has typically been decided as a result of a referendum following public debate. Previous literature has also found a causal effect of the act of recommitting to differentiated integration, as the renewal of the Danish opt-out from justice and home affairs in 2015 led to stronger trust in EU democracy among Danish Eurosceptics [@schraff_does_2020]. The fact that citizens of the region have responded positive to renewals of opt-outs

Both the way that differentiated integration has been framed and the particular national situation of the Scandinavian countries makes it likely that there will be an even stronger positive reaction to differentiation among Scandinavian Eurosceptics: Because their countries are both relatively small and highly dependent on European market access, Eurosceptic Scandinavians could be more likely to see differentiation as a viable second-best alternative to the European disintegration that they otherwise favour. This effect is likely to be compounded by how Scandinavian elites have framed differentiated integration as a tool for greater sovereignty [@adler-nissen_opting_2014]. I thus hypothesize:

@@

To analyse the regionally specific effects of differentiation, I create a dummy for Scandinavian citizenship. I code all Swedish and Danish citizens with a 1 on the Scandinavian dummy. These are used to investigate the interaction effects assumed by *H2*. I also create a dummy for Euroscepticism, coding those who answer that they would vote yes to leaving the European Union in a hypothetical referendum with a 1 and those answering no with a 0. While this only measures what @de_vries_euroscepticism_2018 terms "exit scepticism", an alternative operationalization that uses a measure of satisfaction with EU democracy yields substantively the same results (see Robustness checks).

@@

```{r}
library(modelsummary)
library(flextable)
library(dplyr)
table_format <- ifelse(is_docx, "huxtable", 'default')

#desc_tab <- modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                 #title = "Descriptive statistics", output = "flextable")
#desc_tab %>%
  #autofit()

modelsummary::datasummary_skim(data_descriptives, fmt = 3, histogram = FALSE, 
                 title = "Descriptive statistics", output = "latex")
```

@@

A further limitation, general to all studies using observational survey data, is the potential for omitted variable bias. Here omitted variable bias is particularly likely to stem from the absence of data on the respondents' knowledge of the EU and education [@karp_know_2003; @marquart_knowing_2019]. I use sensitivity analysis [@cinelli_making_2020] to tentatively quantify the size of this bias. I show the results of these analyses in the section Robustness checks. Lastly, the observational data found in this survey cannot be used to make counter-factual claims, as these are discussed in for instance @pearl_causal_2010.

@@

```{r }
main_mod = list()
main_mod$"Euroscepticism" <- lm(opt_out_supp ~ article_disc_log + market_europe + alternative_econ + gov_trust + left_right + left_right_sqr +
                                                         eurosceptic + gender + income_scale + age + excl_identity + article_disc_log*eurosceptic + factor(country), weights = EUI_final_DI_disc$weight, 
                                                         data = EUI_final_DI_disc)
coef_map_eurosceptic <- c('(Intercept)' = 'Constant',
                          'article_disc_log:eurosceptic1' = 'Article-years (logged) X Euroscepticism',
                          "article_disc_log:nordic_dummy1" = "Article-years (logged) X Scandinavia", 
                          "article_disc_log:CEE_dummy1" = "Article-years (logged) X CEE",
                          "nordic_dummy1" = "Scandinavia",
                          "CEE_dummy1" = "Central and Eastern Europe",
                          'article_disc_log' = 'Article-years (logged)',
                          'market_europe' = 'Liberal economic values',
                          'alternative econ' = 'Perception of economy',
                          'gov_trust' = 'Support national government',
                          'left_right' = 'Left-right',
                          'left_right_sqr' = 'Left-right (sqr)',
                          'eurosceptic1' = 'Eurosceptic',
                          'gender' = 'Gender',
                          'age' = 'Age',
                          'income_scale' = 'Income',
                          'excl_identity1' = 'Exclusively national identity')

#analysis1_tab <- modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "flextable", 
 #            stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI. Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))

#analysis1_tab %>%
 # autofit()

modelsummary(main_mod, fmt = 3, coef_map = coef_map_eurosceptic, output = "latex", 
 stars = TRUE, title = "Model showing interaction between Euroscepticism and exposure to DI. Fixed country effects and SEs clustered at country level.", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

Both table 2 and figure 1 show that exposure to differentiated integration is associated with greater support for differentiated integration, and that this effect is particularly prominent among Eurosceptics.

@@

A reasonable counter-argument is that governments are more likely to opt for differentiated integration where underlying popular scepticism of integration may make citizens more likely to accept the idea of differentiated integration. Figure A5 offers some support for this assumption, as there are particularly high levels of differentiation in for instance Sweden and Denmark. These are countries where opt-outs from integration have been deployed as a response to broad Euroscepticism [@leruth_operationalizing_2015; @adler-nissen_opting_2014]. However, citizens of countries as diverse as Lithuania, Hungary, Italy and the Netherlands have also been exposed to differentiated integration. It is thus not a phenomenon restricted to the countries found in what @bolstad_dynamics_2015 identify as a Eurosceptic periphery. This suggests that the correlations do not solely reflect an endogenous relationship between Euroscepticism and differentiation.

@@

#analysis2_tab <- modelsummary(main_regional, fmt = 3, output = "flextable", coef_map = coef_map_regions, 
 #            stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
  #           vcov = ~country, gof_map = c("nobs", "adj.r.squared"))

#analysis2_tab %>%
 # autofit()

modelsummary(main_regional, fmt = 3, output = "latex", coef_map = coef_map_regions, 
     stars = TRUE, title = "Model showing regional interactions (Fixed effects omitted due to collinearity). SEs clustered at country level.", 
            vcov = ~country, gof_map = c("nobs", "adj.r.squared"))
```

@@

```{r, fig.pos="H", fig.cap="Variations in support for differentiation as a function of exposure to DI between Scandinavians and non-Scandinavians. 95% country-clustered CIs. Fixed effects omitted due to multicollinearity.", fig.height=4}
library(patchwork)

@@

#nordic_analysis_tab <- modelsummary::modelsummary(region_mod, fmt = 3, output = "flextable", coef_map = coef_map_regions, 
 #            stars = TRUE, title = "Models showing interactions of Euroscepticism and region. Fixed country effects and SEs clustered at country-level", vcov = ~country, gof_map = c("nobs", "adj.r.squared"))

#nordic_analysis_tab %>%
 # autofit()

@@

I use an approach outlined in @cinelli_making_2020 to test the sensitivity of several models to omitted variable bias (see figures A1-3). The test quantifies the variance that a confounding variable would need to account for to shrink the effect of exposure to differentiated integration to zero. A test of the sensitivity of the interaction between Euroscepticism and exposure to differentiation in Scandinavia (figure A3) shows that a variable explaining 50% of the remaining variance of support for constitutional differentiation would need to explain less than 10% of the remaining variance in real exposure to differentiation to nullify the effect of the variable. One important omitted variable is knowledge of the EU [@leuffen_public_2022]. While knowledge of the EU is unlikely to be strongly connected with actual levels of differentiation, its strong connection to support for differentiation makes its omission particularly problematic. This shows that omitted variable bias is a real concern for this study. Future studies should mitigate this by explicitly measuring knowledge of EU politics, to allow researchers to include a control similar to that found in other EU public opinion research.

I also use multiple imputation [@buuren_mice_2011] to test how sensitive the results are to bias caused by potentially systematic data missingness. By pooling the estimates and standard errors from 25 data imputations according to "Rubin's rules" [@rubin_2004_multiple] I better account for the uncertainty associated with imputing missing values. This is particularly beneficial when conducting significance test. The effects of the multiple imputation suggests that systematic missingness may be an issue: Table A2 shows that the effect of the interaction between exposure to differentiation and Euroscepticism loses its significance when the data are imputed. Similarly, we no longer find a significant interaction effect between exposure and Euroscepticism in Scandinavia (table A4). However, the interaction effect of exposure and Scandinavian citizenship appears strongly robust to imputation.

I also test whether the substantial overlap between Scandinavian Eurosceptics and non-Eurosceptics is driven by the low variability, with a standard deviation of `r round(sd(nordic_data$article_disc_log), digits = 2)`, of the exposure to differentiated integration found in the two Scandinavian countries. I do this by testing a three-way interaction between Scandinavian citizenship, exposure to differentiated integration and Euroscepticism in the full data. Appendix table A7 shows that the results are substantially similar to the interaction effect of Scandinavian citizenship and Euroscepticism shown in table 4. The results thus suggest that the estimates are not unduly biased by the variability found in the Scandinavian data.

To test the scope of the established effect, I restrict my analysis to differentiated integration among the policies @genschel_more_2016 term "core state powers", state powers that are essential to state functioning. A reasonable assumption might be that citizens could be especially inclined towards more differentiation if exposed to differentiations in the field of core state power integration, as these are areas where citizens might find it important to retain national sovereignty [@hooghe_postfunctionalist_2009; @de_wilde_can_2012]. My results, visualized by predicted values (figure A7), suggest that policies in question matter for the effect of differentiation: While the effect of exposure to differentiated core state power integration seems largely flat among supporters of EU membership, it increases with exposure to differentiation among Eurosceptics. However, the effect seems largely comparable to those visualized in figure 1. While the reason for this may be that much differentiated integration has happened in the field of core state powers [@schimmelfennig_ever_2020], it also suggests that the effects are robust to alternative operationalizations of the measure of exposure to differentiated integration.

@@

This paper establishes, by combining data surveying citizen preferences for differentiated integration with information about their countries' actual exposure to differentiation, that there may be a strong association between support for differentiated integration and past exposure to it. The reason may be that those accustomed to differentiated integration have become more favourably inclined towards it because they have learnt from experience how it can combine membership of the EU with greater exercise of national autonomy. Importantly, the effect appears moderated by whether one opposes or supports EU membership. It also seems as if exposure to frames depicting differentiated integration as a source of greater autonomy could be an important driver of support for differentiated integration

The fact that Eurosceptics become more favourably inclined to differentiated integration the more they are exposed to it, even if they primarily want an exit from the European Union, suggests that differentiated integration may be seen as an acceptable "second-best" alternative to an EU exit even among those considering themselves opponents of membership. This strengthens the assumption that differentiated integration, as is pointed to in previous literature [@bellamy_republican_2019; @schimmelfennig_ever_2020], can mitigate the contestation of European integration brought by the EU's expansion into more important areas.
